\myheading{Modulo inverse, part II}

A very simple function:

\begin{lstlisting}
int f(int a)
{
        return a/9;
};
\end{lstlisting}

If compiled by non-optimizing GCC 4.4.1...

\lstinputlisting[caption=Non-optimizing GCC 4.4.1]{modulo/11_2_gcc.asm}

And it can be rewritten back to pure C like that:

\begin{lstlisting}[style=customc]
#include <stdint.h>

uint32_t divide_by_9 (uint32_t a)
{
        return ((uint64_t)a * (uint64_t)954437177) >> 33; // 954437177 = 0x38e38e39
};
\end{lstlisting}

This function still works, without division operation. How?

From school-level mathematics, we can remember that division by 9 can be replaced by multiplication by $\frac{1}{9}$.
In fact, sometimes compilers do so for floating-point arithmetics, for example, \texttt{FDIV} instruction in x86 code
can be replaced by \texttt{FMUL}.
At least MSVC 6.0 will replace division by 9 by multiplication by $0.111111...$ and sometimes it's hard to be sure,
what operation was in the original source code.

But when we operate over integer values and integer CPU registers, we can't use fractions.
However, we can rework fraction like that:

% FIXME: equation size
\begin{center}
$result = \frac{x}{9} = x \cdot \frac{1}{9} = x \cdot \frac{1 \cdot MagicNumber}{9 \cdot MagicNumber}$
\end{center}

Given the fact that division by $2^n$ is very fast (using shifts), we now should find that $MagicNumber$,
for which the following
equation will be true: $2^n = 9 \cdot MagicNumber$.

Division by $2^{32}$ is somewhat hidden: lower 32-bit of product in EAX is not used (dropped), only higher 32-bit of
product (in EDX) is used and then shifted by additional 1 bit.

In other words, the assembly code we have just seen multiplicates by {\Large $\frac{954437177}{2^{32+1}}$},
or divides by {\Large $\frac{2^{32+1}}{954437177}$}.
To find a divisor we just have to divide numerator by denominator.
Using Wolfram Alpha, we can get 8.99999999.... as result (which is close to 9).

% TODO ref to https://yurichev.com/blog/signed_division_using_shifts/

% FIXME:
% Read more about it in \InSqBrackets{\HenryWarren 10-3}.

Many people miss ``hidden'' division by $2^{32}$ or $2^{64}$,
when lower 32-bit part (or 64-bit part) of product is not used.
This is why this division by multiplication is difficult to understand to many people...

