\myheading{Reversible linear congruential generator}

\ac{LCG} \ac{PRNG} is very simple: just multiply seed by some value, add another one and here is a new random number.
Here is how it is implemented in MSVC (the source code is not original one and is reconstructed by me):</p>

\begin{lstlisting}[style=customc]
uint32_t state;

uint32_t rand()
{
	state=state*214013+2531011;
	return (state>>16)&0x7FFF;
};
\end{lstlisting}

The last bit shift is attempt to compensate LCG weakness and we may ignore it so far.
Will it be possible to make an inverse function to rand(), which can reverse state back?
First, let's try to think, what would make this possible? Well, if state internal variable would be some kind of BigInt or BigNum container which can
store infinitely big numbers, then, although state is increasing rapidly, it would be possible to reverse the process.
But \textit{state} isn't BigInt/BigNum, it's 32-bit variable, and summing operation is easily reversible on it (just subtract 2531011 at each step).
As we may know now, multiplication is also reversible: just multiply the state by modular multiplicative inverse of 214013!

\begin{lstlisting}[style=customc]
#include <stdio.h>
#include <stdint.h>

uint32_t state;

void next_state()
{
	state=state*214013+2531011;
};

void prev_state()
{
	state=state-2531011; // reverse summing operation
	state=state*3115528533; // reverse multiply operation. 3115528533 is modular inverse of 214013 in 2^32.
};

int main()[style=customc]
{
	state=12345;
	
	printf ("state=%d\n", state);
	next_state();
	printf ("state=%d\n", state);
	next_state();
	printf ("state=%d\n", state);
	next_state();
	printf ("state=%d\n", state);

	prev_state();
	printf ("state=%d\n", state);
	prev_state();
	printf ("state=%d\n", state);
	prev_state();
	printf ("state=%d\n", state);
};
\end{lstlisting}

Wow, that works!

\begin{lstlisting}
state=12345
state=-1650445800
state=1255958651
state=-456978094
state=1255958651
state=-1650445800
state=12345
\end{lstlisting}

It's hard to find a real-world application of reversible LCG, but this could be the one:
a media player with forward/backward buttons.
Once shuffle is clicked, random number is generated (number of item to be played).
User clicks forward: get a new random number by calculating the next state.
User clicks backward: get it by calculating the previous state.
Thus, a user could navigate through some "virtual" (but consistent) playlist, which is even not present in media player's memory!

